\systemEnding: Bootstrapping Domain-Specific Languages for Neurally-Guided Bayesian Program Learning

### 2019-04-19 archive.tex
even more writing

Because solutions to many kinds of problems can often be described as
some kind of
program~\cite{lake2015human,Ullman2012,DBLP:journals/cogsr/SchmidK11},
our model solves a specific problem by synthesizing a program solving
it. It gradually grows its explicit domain-specific knowledge for a
class of problems by assembling a library of code containing concepts
useful for the domain, and by training a neural network to quickly
infer, for a specific problem, what code in its library is likely to
solve it. We think of these two learned components as analogous to the
explicit concepts and implicit procedural skill that human experts
develop.  The model is an instance of what in the machine learning
community is called a wake-sleep algorithm, where for us, `waking'
corresponds to solving problems, while a pair of `sleep' cycles
correspond to improving the library of explicit concepts and training
the neural net to improve implicit procedural knowledge of how to
write code.

@@

Our model iteratively creates new library routines that build on
concepts learned earlier in its learning trajectory, growing a library with nested
hierarchies of code.  We think of this cumulative nesting of
abstractions as a variety of deep representation
learning~\cite{lecun2015deep}.  Figure~\ref{exampleDSL} diagrams a
subset of these learned networks (the library). For example, the model
learns to sort sequences of numbers by invoking a library component 4
layers deep, or draws the leftmost pairs of images in
Figure~\ref{exampleDSL} using a depth-3 component.  For this reason we
refer to our approach as an instance of `deep program learning'.

@@

Could we also learn the basics of programming, including these recursive sequence manipulation primitives?
This goal is relevant to a long-standing dream within the program induction community to ``learn from scratch'': starting with a \emph{minimal} Turing-complete programming language,
and then learning to solve a wide swath of
induction problems~\cite{solomonoff1964formal,schmidhuber2004optimal,hutter2004universal,solomonoff1989system}.
All existing systems,
including ours,
fall far short of this dream,
and it is unclear (and we believe unlikely)
that this dream could ever be fully realized.
How far can we push in this direction?
``Learning from scratch'' is subjective, but a reasonable
starting point is the set of primitives provided in 1959
Lisp~\cite{mccarthy1960recursive}: these include
conditionals, recursion, arithmetic, and the 
list operators \code{cons}, \code{car}, \code{cdr}, and \code{nil}.
A  basic first goal is to start with
these primitives,
and then recover a DSL that
more closely resembles modern functional languages like Haskell and OCaml.
Recall (Sec.~\ref{sequences})
that we initially provided our system with functional programming routines like
\code{map} and \code{fold}.

@@

\subsection{Analyzing \system across domains}\label{quantitative}
To evaluate the relative importance of library learning and recognition
model training, we evaluate on held-out testing tasks for each of our
domains, measuring both how many tasks are solved and how long it
takes to solve them across successive wake/sleep iterations
(Fig.~\ref{learningCurves}).  We always solve more held-out tasks --
and generally solve them in less time -- with both components
combined, suggesting a synergy between library learning and
recognition model training. Examining the libraries learned with and without
recognition models reveals a difference in the typical \emph{depth}
of learned subroutines, and this average depth correlates
with the number of solved testing tasks ($r=0.71$, $p < 10^{-8}$).
In contrast, the correlation between the \emph{number} of learned subroutines and number of solved tasks is weaker and not statistically significant ($r=0.20$, $p > 0.05$),
suggesting that the recognition model bootstraps
``better'' libraries, where ``better'' is at least partly explained by the depth
of the learned representation. Figure~\ref{depthMontage} shows typical `deep' libraries learned by
\system and contrasts this depth with \% solved tasks, both with and without the recognition model.
\begin{figure}
  \begin{tabular}{ccc}
    \begin{tabular}{c}
      \includegraphics[width = 5cm]{figures/learningCurves/text_hits.png}\\
      \includegraphics[width = 5cm]{figures/learningCurves/text_time.png}
    \end{tabular}&

>>>>

Because solutions to many kinds of problems can often be described as
some kind of
program~\cite{lake2015human,Ullman2012,DBLP:journals/cogsr/SchmidK11},
our model approaches a specific problem by searching for a program solving
it. It gradually grows its explicit domain-specific knowledge for a
class of problems by assembling a library of code containing concepts
useful for the domain, and by training a neural network to quickly
infer, for a specific problem, what code in its library is likely to
solve it. We think of these two learned components as analogous to the
explicit concepts and implicit procedural skill that human experts
develop.  The model is an instance of what in the machine learning
community is called a wake-sleep algorithm, where for us, `waking'
corresponds to solving problems, while a pair of `sleep' cycles
correspond to improving the library of explicit concepts and training
the neural net to improve implicit procedural knowledge of how to
write code.

@@

Our model iteratively creates new library routines that build on
concepts acquired earlier in its learning trajectory, growing a library with nested
hierarchies of code.  We think of this cumulative nesting of
abstractions as a variety of deep representation
learning~\cite{lecun2015deep}.  Figure~\ref{exampleDSL} diagrams a
subset of these learned networks (the library). For example, the model
learns to sort sequences of numbers by invoking a library component 4
layers deep, or draws the leftmost pairs of images in
Figure~\ref{exampleDSL} using a depth-3 component.  For this reason we
refer to our approach as an instance of `deep program learning'.

@@

Could we also learn the basics of programming, including these recursive sequence manipulation primitives?
This goal is relevant to a long-standing dream within the program induction community to ``learn from scratch'': starting with a \emph{minimal} Turing-complete programming language,
and then learning to solve a wide swath of
induction problems~\cite{solomonoff1964formal,schmidhuber2004optimal,hutter2004universal,solomonoff1989system}.
All existing systems,
including ours,
fall far short of this dream,
and it is unclear (and we believe unlikely)
that this dream could ever be fully realized.
How far can we push in this direction?
``Learning from scratch'' is subjective, but a reasonable
starting point is the set of primitives provided in 1959
Lisp~\cite{mccarthy1960recursive}: these include
conditionals, recursion, arithmetic, and the 
list operators \code{cons}, \code{car}, \code{cdr}, and \code{nil}.
A  basic first goal is to start with
these primitives,
and then recover a DSL that
more closely resembles modern functional languages like Haskell and OCaml.

@@

\subsection{Analyzing \system across domains}\label{quantitative}
To evaluate the relative importance of library learning and recognition
model training, we evaluate on held-out testing tasks for each of our
domains, measuring both how many tasks are solved and how long it
takes to solve them across successive wake/sleep iterations
(Fig.~\ref{learningCurves}).  We always solve more held-out tasks --
and generally solve them in less time -- with both components
combined, suggesting a synergy between these forms of declarative and procedural knowledge.
Examining the libraries learned with and without
recognition models reveals a difference in the typical \emph{depth}
of learned subroutines, and this average depth correlates
with the number of solved testing tasks ($r=0.71$, $p < 10^{-8}$).
In contrast, the correlation between the \emph{number} of learned subroutines and number of solved tasks is weaker and not statistically significant ($r=0.20$, $p > 0.05$),
suggesting that the recognition model bootstraps
``better'' libraries, where ``better'' is at least partly explained by the depth
of the learned representation. Figure~\ref{depthMontage} shows typical `deep' libraries learned by
\system and contrasts this depth with \% solved tasks, both with and without the recognition model.
\begin{figure}
  \begin{tabular}{ccc}
    \begin{tabular}{c}
      \includegraphics[width = 5cm]{figures/learningCurves/text_hits.png}\\
      \includegraphics[width = 5cm]{figures/learningCurves/text_time.png}
    \end{tabular}&
